A linearization procedure for constrained multibody systems

### 2012-04-21 PetersonGedeHubbard2012.tex
Major rearrangment of paper sections.

\documentclass{svjour3}                     % onecolumn (standard format)
\RequirePackage{fix-cm}

\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{bm}

@@

\section{Introduction}
\label{intro}

The goal of the following calculations is to concretely establish the first
order relationships (about an arbitrary equilibrium point) between an arbitrary
choice of independent coordinates and speeds, and the time-derivatives of all
coordinates and speeds (both independent and dependent).  This is necessary for
tasks which require a linear state space formulation.  Common examples include
stability analysis and controller and estimator design (including control or
measurement of a dependent quantity).  The formalism presented here is generic
enough to cover most examples of time-varying constrained multibody systems
with arbitrary external inputs and arbitrary specified quantities.  This
procedure has been created for systems whose equations of motion have been
derived with Kane's Method.  While this is not a strict requirement, systems
which can be completely described by a set of ODEs is (i.e., no DAEs).

The complete configuration of the system is further assumed to be described by
generalized coordinates $\bm{q}\in\bm{R}^n$ and its velocity completely
described by generalized speeds $\bm{u}\in\bm{R}^o$, with $n \ne o$ in general.
All coordinates and speeds, not only the independent ones, are represented by
$\bm{q}$ and $\bm{u}$.  The system is assumed to be governed by the
relationships described in Table \ref{table:assumptions}.  The system is
described by $l + m + n + o$ equations which relate $2n + 2o + s + 1$
quantities, which implies that $n - l + o - m + s + 1$ of these quantities are
independent.  This paper discusses the first order relationship between these
independent quantities ($n-l$ independent coordinates, $o-m$ independent
speeds, $s$ control inputs, and time $t$) and the time derivatives of the $n$
coordinates and $o$ speeds.
\begin{table}[htbp]
  \centering
  \caption{Constrained multibody system governing definitions and equations}
  \label{table:assumptions}
  \begin{tabular}[c]{l l l}
    Quantity & Shape & Description\\
    \hline
    $\bm{q},\bm{\dot{q}}$ & $\mathbf{R}^n$ & Coordinates and their time
    derivatives\\
    $\bm{u}, \bm{\dot{u}}$ & $\mathbf{R}^o$ & Speeds and their time derivatives\\
    $\bm{r}$ & $\mathbf{R}^s$ & Exogenous inputs \\
  \end{tabular}
  \begin{tabular}[c]{r @{ $=$ } l l}
    \multicolumn{3}{c}{ } \\
    \multicolumn{2}{c}{Equation} & Description \\
    \hline
    $\bm{f}_{c}(\bm{q}, t)$ & $\bm{0}_{l \times 1}$ & Configuration constraints \\
    $\bm{f}_{v}(\bm{q}, \bm{u}, t)$ & $\bm{0}_{m \times 1}$ & Velocity constraints \\
    $\bm{f}_{a}(\bm{q}, \bm{\dot{q}}, \bm{u}, \bm{\dot{u}}, t)$ & $\bm{0}_{m
    \times 1}$ & Acceleration constraints \\
    $\bm{f}_{0}(\bm{q}, \bm{\dot{q}}, t) + \bm{f}_{1}(\bm{q}, \bm{u}, t)$ &
    $\bm{0}_{n \times 1}$ & Kinematic differential equations \\
    $\bm{f}_{2}(\bm{q}, \bm{\dot{u}}, t) + \bm{f}_{3}(\bm{q}, \bm{\dot{q}},
    \bm{u}, \bm{r}, t)$ & $\bm{0}_{(o - m) \times 1}$ & Dynamic differential equations
  \end{tabular}
\end{table}

The first three equations in Table \ref{table:assumptions} represent
constraints derived purely from kinematic considerations at the configuration,
velocity, and acceleration levels, respectively.  The velocity constraints may
be nonholonomic or time differentiated holonomic.  The acceleration constraints
can be time differentiated velocity constraints or kinematic acceleration
constraints.  The fourth equation represents kinematic differential equations
which relate $\bm{\dot{q}}$ to $\bm{u}$ and are linear in both these
quantities.  The fifth equation represents the constraint-free nonholonomic
dynamic differential equations~\cite{Kane1985}, and will necessarily be linear
in $\bm{\dot{u}}$ and the inputs $\bm{r}$, but generally nonlinear in all other
quantities.  This formalism is exceedingly generic; in practice, the structure
of the functions in Table \ref{table:assumptions} can be leveraged to simplify
the following calculations.  This will be discussed subsequently.

Trajectories of the system exist on a $p \triangleq n - l + o - m$ dimensional
manifold embedded in a $n + o$ dimensional space, though conserved quantities
(e.g., total mechanical energy) may restrict trajectories to a lower
dimensional manifold.  In general this cannot be assumed, however, especially
in the presence of exogenous inputs $\bm{r}$ which may change the total
mechanical energy of the system.  While more than $p$ of the independent state
variables may be of interest (i.e., dependent coordinates or speeds may be of
interest in themselves), only $p$ quantities are truly independent.  Which of
these $p$ coordinates and speeds should be chosen as independent out of the
possible $n + o$ may be obvious for small systems where there is intuition
about system behavior, but in general the optimal choice depends on how the
configuration $\bm{q}$ affects the gradients of $f_c$ and $f_v$.  On the other
hand, in systems with cyclic (ignorable) coordinates, fewer than $p$ state
variables may be of interest; one example is the Whipple bicycle model where
$p=10$ but 5 of these state variables are cyclic coordinates (Cartesian
coordinates of rear wheel contact point, heading, and both wheel angles) and
may only be of secondary interest.

@@

Equating the right hand sides of equations (\ref{eq:f0_plus_f1_expansion}),
(\ref{eq:acceleration_expansion}),
and (\ref{eq:f2_plus_f3_expansion}) to zero (as per Table
\ref{table:assumptions}), and introducing the following definitions
\begin{align}
    \label{eq:quant_to_compute}
    \begin{array}{llcll}
\tilde{M}_{qq}  &\triangleq \nabla_{\bm{\dot{q}}}\bm{f}_0 & \quad &
\tilde{M}_{uqc} &\triangleq \nabla_{\bm{\dot{q}}}\bm{f}_a \\
\tilde{M}_{uuc} &\triangleq \nabla_{\bm{\dot{u}}}\bm{f}_a & \quad &
\tilde{M}_{uqd} &\triangleq \nabla_{\bm{\dot{q}}}\bm{f}_2 \\
\tilde{M}_{uud} &\triangleq \nabla_{\bm{\dot{u}}}\bm{f}_2 & \quad &
\tilde{A}_{qq}  &\triangleq -\nabla_{\bm{q}}(\bm{f}_0 + \bm{f}_1) \\
\tilde{A}_{qu}  &\triangleq -\nabla_{\bm{u}}\bm{f}_1 & \quad &
\tilde{A}_{uqc} &\triangleq - \nabla_{\bm{q}} \bm{f}_a \\
\tilde{A}_{uuc} &\triangleq - \nabla_{\bm{u}} \bm{f}_a & \quad &
\tilde{A}_{uqd} &\triangleq - \nabla_{\bm{q}} (\bm{f}_2 + \bm{f}_3) \\
\tilde{A}_{uud} &\triangleq - \nabla_{\bm{u}} \bm{f}_3 & \quad &
\tilde{B}_{u}   &\triangleq -\nabla_{\bm{r}}\bm{f}_{3}
\end{array}
\end{align}
enables the unconstrained linear state space equations to be written as
\begin{align}
  \label{eq:state_space_unconstrained}
  \left[
    \begin{array}{cc}
      \tilde{M}_{qq} & \bm{0}_{n \times o} \\
      \tilde{M}_{uqc} & \tilde{M}_{uuc} \\
      \tilde{M}_{uqd} & \tilde{M}_{uud}
    \end{array}
    \right]
    \left[
      \begin{array}{c}
        \delta \bm{\dot{q}} \\
        \delta \bm{\dot{u}}
      \end{array}
    \right]
   &=
   \left[
     \begin{array}{cc}
       \tilde{A}_{qq} & \tilde{A}_{qu} \\
       \tilde{A}_{uqc} & \tilde{A}_{uuc} \\
       \tilde{A}_{uqd} & \tilde{A}_{uud}
     \end{array}
   \right]
    \left[
      \begin{array}{c}
        \delta \bm{q} \\
        \delta \bm{u}
      \end{array}
    \right]
    +
    \left[
      \begin{array}{c}
        \bm{0}_{(n + m) \times s} \\
        \tilde{B}_{u}
      \end{array}
    \right]
    \delta \bm{r}
\end{align}
Equation (\ref{eq:state_space_unconstrained}) has a state space of dimension $n
+ o$, yet only $p = n - l + o - m$ of these quantities are independent.
To address this issue, a particular set of independent
coordinates and speeds must be selected. To this end, consider the following
partitioning of the generalized coordinates and generalized speeds:
\begin{equation*}
  \tilde{\bm{q}} \triangleq \left[\begin{array}{cc}\bm{q}_{i} &
      \bm{q}_{d}\end{array}\right]^{T} =  P_{q}^{-1} \bm{q}
      \qquad\qquad
  \tilde{\bm{u}} \triangleq \left[\begin{array}{cc}\bm{u}_{i} &
      \bm{u}_{d}\end{array}\right]^{T} =  P_{u}^{-1} \bm{u}
\end{equation*}
where $P_q \in \mathbf{R}^{n \times n}$ and $P_u \in \mathbf{R}^{o \times o}$
are invertible permutation matrices which map an ordering which has the
independent quantities ($\bm{q}_{i}\in\mathbf{R}^{n-l},\,
\bm{u}_{i}\in\mathbf{R}^{o-m})$ first, followed by the dependent quantities
($\bm{q}_{d}\in\mathbf{R}^{l},\, \bm{u}_{d}\in\mathbf{R}^{m}$) to the
original ordering of the coordinates and speeds.  We use the notation $P_{qi}$
and $P_{qd}$ to denote the first $n-l$ and last $l$ columns of $P_q$,
respectively; similarly, $P_{ui}$ is the first $o-m$ columns of $P_{u}$ while
$P_{ud}$ is the last $m$ columns of $P_u$.

By assumption, equation
(\ref{eq:configuration_expansion}) is zero; making use of $P_q$, we have
\begin{align}
  \bm{0} &= \nabla_{\bm{q}}\bm{f}_{c} P_{q} \delta \bm{\tilde{q}} \notag \\
   &= \nabla_{\bm{q}}\bm{f}_{c} P_{qi} \delta \bm{q_i} +
  \nabla_{\bm{q}}\bm{f}_{c} P_{qd} \delta \bm{q_d}\notag\\
  \implies \delta \bm{q}_d &= -(\nabla_{\bm{q}}\bm{f}_{c} P_{qd})^{-1}
  (\nabla_{\bm{q}}\bm{f}_{c} P_{qi}) \delta \bm{q}_i \notag\\
  \implies \delta \bm{q} &= \left[ I_{n \times n} - P_{qd}(\nabla_{\bm{q}}
    \bm{f}_{c} P_{qd})^{-1} \nabla_{\bm{q}} \bm{f}_{c} \right] P_{qi} \delta
    \bm{q}_i
\end{align}
We define
\begin{equation}
  \label{eq:C_0}
  C_0 \triangleq \left[ I_{n \times n} - P_{qd}(\nabla_{\bm{q}}
    \bm{f}_{c} P_{qd})^{-1} \nabla_{\bm{q}} \bm{f}_{c} \right] P_{qi}
\end{equation}
which enables us to write
\begin{align}
  \label{eq:delta_q}
  \delta \bm{q} &= C_0 \delta \bm{q}_i
\end{align}

Applying the same approach to (\ref{eq:velocity_expansion}), we obtain
\begin{align}
  \bm{0} &= \nabla_{\bm{q}}\bm{f}_{v} \delta \bm{q} +
  \nabla_{\bm{u}}\bm{f}_{v} \delta \bm{u}\notag\\
  &= \nabla_{\bm{q}} \bm{f}_{v}
\delta \bm{q} + \nabla_{\bm{u}} \bm{f}_{v} P_{ui} \delta \bm{u}_i +
\nabla_{\bm{u}} \bm{f}_{v} P_{ud} \delta \bm{u}_d \notag\\
\implies \delta \bm{u}_d &= -\left(\nabla_{\bm{u}} \bm{f}_{v}
P_{ud}\right)^{-1}\left[\nabla_{\bm{q}}\bm{f}_{v} \delta\bm{q} +
  \nabla_{\bm{u}} \bm{f}_{v} P_{ui} \delta \bm{u}_i \right]\notag\\
  \implies \delta \bm{u} &= -P_{ud}(\nabla_{\bm{u}} \bm{f}_{v} P_{ud})^{-1}
  \nabla_{\bm{q}} \bm{f}_{v} \delta \bm{q}\notag\\
  &+ \left[I - P_{ud} (\nabla_{\bm{u}}\bm{f}_{v} P_{ud})^{-1} \nabla_{\bm{u}}
    \bm{f}_{v} \right] P_{ui} \delta \bm{u}_i
\end{align}
We define
\begin{align}
  \label{eq:C_1}
  C_1 &\triangleq -P_{ud}(\nabla_{\bm{u}} \bm{f}_{v} P_{ud})^{-1}
  \nabla_{\bm{q}} \bm{f}_{v} \\
  \label{eq:C_2}
  C_2 &\triangleq \left[I - P_{ud} (\nabla_{\bm{u}}\bm{f}_{v} P_{ud})^{-1} \nabla_{\bm{u}}
    \bm{f}_{v} \right] P_{ui}
\end{align}
which enables us to write
\begin{align}
  \label{eq:delta_u}
  \delta \bm{u} &= C_1 \delta \bm{q} + C_2 \delta \bm{u}_i\notag\\
  &= C_1 C_0 \delta \bm{q}_i + C_2 \delta \bm{u}_i
\end{align}

>>>>

\documentclass[smallcondensed]{svjour3}                     % onecolumn (standard format)
\RequirePackage{fix-cm}

\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{bm}
%\usepackage[polutonikogreek]{babel}

@@

\section{Conclusions}
\label{sec:conclusions}
We have provided a procedure to obtain linear state space formulation for
constrained multibody systems whose nonlinear dynamics are described only by
ordinary differential equations. With this state space formulation, you can
perform stability analysis, control system design, and other linear analyses.

In some sense this work is simply applying the chain rule through matrix
operations in a way that respects system constraints. An advanced C.A.S. system
which possesses the ability to represent dependent quantities (as functions of
independent quantities) could perform the chain rule directly, thereby
eliminating the need for matrix operations. We leave an open question: how does
computational performance depend on implementation details and problem size?

%In summary, the procedure is:
%\begin{enumerate}
%    \item Describe the system in the form of the equations shown in Table
%        \ref{table:assumptions}.
%    \item Determine a point of linearization ($\bm{q}^*$, $\dot{\bm{q}}^*$,
%      $\bm{u}^*$, $\dot{\bm{u}}^*$, and $\bm{r}^*$) which satisfies the
%      equations in Table \ref{table:assumptions}.
%    \item Compute the quantities in equation (\ref{eq:quant_to_compute}).
%    \item Identify independent coordinates and speeds; form equations (\ref{eq:C_0}), (\ref{eq:C_1}), and
%        (\ref{eq:C_2}).
%    \item Form equation (\ref{eq:state_space_constrained}), and if needed,
%      equations (\ref{eq:A_prime})-(\ref{eq:B}).
%\end{enumerate}

\section{Introduction}
\label{sec:intro}
Multibody systems, rigid or flexible, and particles often have constraints
which limit how they can move. In this paper, we will discuss 3 types of
constraints: configuration, velocity, and acceleration. Configuration
constraints limit the location or orientation of parts of the system, relative
to the external world or other parts of the system. Velocity constraints limit
the speeds in which the configuration can change, either from configuration
constraints (through time differentiation) or independent application. In this
paper, acceleration constraints refer to time differentiated velocity
constraints. Examples of multibody systems with constraints are closed
kinematic loops or objects rolling without slip.

For these systems, a linear form of the equations of motion will be necessary
for many different analyses.  Stability analysis and control system design are
two possible scenarios.  Linearization of these constrained systems is not
straightforward and presents significant challenges. Simply computing the
jacobian matrix of the differential equations will give incorrect results, with
no immediate indication that the results are wrong.

Currently, there are only a few approaches to linearizing constrained systems
which have been presented (citations here). A limitation of these approaches is
that they have been developed for equations of motion found using Lagrange¿s
method, which contain Lagrange multipliers. Kane¿s method (citation here)
allows for formulation of equations of motion for constrained systems in which
no multipliers are present. The dynamics of these systems can be completely
described by ordinary differential equations. This significantly different
formulation requires a different approach to linearization than what the
existing options provide.

The goal of the following paper is to concretely establish the first order
relationships (about an arbitrary equilibrium point) between a selection of
independent coordinates and speeds and the time-derivatives of all of the
coordinates and speeds (both independent and dependent). The formalism
presented here is generic enough to cover most examples of time-varying
constrained multibody systems with arbitrary external inputs and arbitrary
specified quantities.  This procedure was created for systems whose equations
of motion have been derived with Kane¿s Method. Although also applicable to
dynamic system equations formulated using other methods, it is restricted to
systems which can be completely described by a set of ODEs (i.e., no DAEs are
present).

We first present our linearization procedure. Next we will offer some
discussion of the procedure, along with an example. The derivation of the
procedure is provided in the appendix. Additionally, the electronic
supplementary material provides more examples (in more detail) and an
alternative ordering for the independent/dependent quantities.

%The complete configuration of the system is further assumed to be described by
%generalized coordinates $\bm{q}\in\bm{R}^n$ and its velocity completely
%described by generalized speeds $\bm{u}\in\bm{R}^o$, with $n \ne o$ in general.
%All coordinates and speeds, not only the independent ones, are represented by
%$\bm{q}$ and $\bm{u}$.  The system is assumed to be governed by the
%relationships described in Table \ref{table:assumptions}.  The system is
%described by $l + m + n + o$ equations which relate $2n + 2o + s + 1$
%quantities, which implies that $n - l + o - m + s + 1$ of these quantities are
%independent.  This paper discusses the first order relationship between these
%independent quantities ($n-l$ independent coordinates, $o-m$ independent
%speeds, $s$ control inputs, and time $t$) and the time derivatives of the $n$
%coordinates and $o$ speeds.
%
%The first three equations in Table \ref{table:assumptions} represent
%constraints derived purely from kinematic considerations at the configuration,
%velocity, and acceleration levels, respectively.  The velocity constraints may
%be nonholonomic or time differentiated holonomic.  The acceleration constraints
%can be time differentiated velocity constraints or kinematic acceleration
%constraints.  The fourth equation represents kinematic differential equations
%which relate $\bm{\dot{q}}$ to $\bm{u}$ and are linear in both these
%quantities.  The fifth equation represents the constraint-free nonholonomic
%dynamic differential equations~\cite{Kane1985}, and will necessarily be linear
%in $\bm{\dot{u}}$ and the inputs $\bm{r}$, but generally nonlinear in all other
%quantities.  This formalism is exceedingly generic; in practice, the structure
%of the functions in Table \ref{table:assumptions} can be leveraged to simplify
%the following calculations.  This will be discussed subsequently.
%
%Trajectories of the system exist on a $p \triangleq n - l + o - m$ dimensional
%manifold embedded in a $n + o$ dimensional space, though conserved quantities
%(e.g., total mechanical energy) may restrict trajectories to a lower
%dimensional manifold.  In general this cannot be assumed, however, especially
%in the presence of exogenous inputs $\bm{r}$ which may change the total
%mechanical energy of the system.  While more than $p$ of the independent state
%variables may be of interest (i.e., dependent coordinates or speeds may be of
%interest in themselves), only $p$ quantities are truly independent.  Which of
%these $p$ coordinates and speeds should be chosen as independent out of the
%possible $n + o$ may be obvious for small systems where there is intuition
%about system behavior, but in general the optimal choice depends on how the
%configuration $\bm{q}$ affects the gradients of $f_c$ and $f_v$.  On the other
%hand, in systems with cyclic (ignorable) coordinates, fewer than $p$ state
%variables may be of interest; one example is the Whipple bicycle model where
%$p=10$ but 5 of these state variables are cyclic coordinates (Cartesian
%coordinates of rear wheel contact point, heading, and both wheel angles) and
%may only be of secondary interest.

\section{Results}
\label{sec:results}
The linearized equations of motion are
\begin{align}
  \label{eq:state_space_constrained}
  \left[
    \begin{array}{cc}
      \tilde{M}_{qq} & \bm{0}_{n \times o} \\
      \tilde{M}_{uqc} & \tilde{M}_{uuc} \\
      \tilde{M}_{uqd} & \tilde{M}_{uud}
    \end{array}
    \right]
    \left[
      \begin{array}{c}
        \delta \bm{\dot{q}} \\
        \delta \bm{\dot{u}}
      \end{array}
    \right]
   &=
   \left[
     \begin{array}{cc}
       (\tilde{A}_{qq} + \tilde{A}_{qu} C_1 ) C_0 & \tilde{A}_{qu} C_2 \\
       (\tilde{A}_{uqc} + \tilde{A}_{uuc} C_1 ) C_0 & \tilde{A}_{uuc} C_2\\
       (\tilde{A}_{uqd} + \tilde{A}_{uud} C_1 ) C_0 & \tilde{A}_{uud} C_2
     \end{array}
   \right]
    \left[
      \begin{array}{c}
        \delta \bm{q}_i \\
        \delta \bm{u}_i
      \end{array}
    \right]
    +
    \left[
      \begin{array}{c}
        \bm{0}_{(n+m) \times s} \\
        \tilde{B}_{u}
      \end{array}
    \right]
    \delta \bm{r}
\end{align}
Table \ref{table:assumptions} provides the form in which the nonlinear
equations of motion and constraints must be arranged.  Equations
\ref{eq:quant_to_compute} provides the bulk of the terms in equation
\ref{eq:state_space_constrained}; these terms are Jacobian matrices of the
differential equations. Equations \ref{eq:C_0}-\ref{eq:C_2} provide the terms
$C_0$, $C_1$, and $C_2$ which embed the constraints into
\ref{eq:state_space_constrained}. Finally, the permutation matrices in
equations \ref{eq:Pq} and \ref{eq:Pu} provide the means to separate the
independent and dependent speeds.

\begin{table}[htbp]
  \centering
  \caption{Constrained multibody system governing definitions and equations}
  \label{table:assumptions}
  \begin{tabular}[c]{l l l}
    Quantity & Shape & Description\\
    \hline
    $\bm{q},\bm{\dot{q}}$ & $\mathbf{R}^n$ & Coordinates and their time
    derivatives\\
    $\bm{u}, \bm{\dot{u}}$ & $\mathbf{R}^o$ & Speeds and their time derivatives\\
    $\bm{r}$ & $\mathbf{R}^s$ & Exogenous inputs \\
  \end{tabular}
  \begin{tabular}[c]{r @{ $=$ } l l}
    \multicolumn{3}{c}{ } \\
    \multicolumn{2}{c}{Equation} & Description \\
    \hline
    $\bm{f}_{c}(\bm{q}, t)$ & $\bm{0}_{l \times 1}$ & Configuration constraints \\
    $\bm{f}_{v}(\bm{q}, \bm{u}, t)$ & $\bm{0}_{m \times 1}$ & Velocity constraints \\
    $\bm{f}_{a}(\bm{q}, \bm{\dot{q}}, \bm{u}, \bm{\dot{u}}, t)$ & $\bm{0}_{m
    \times 1}$ & Acceleration constraints \\
    $\bm{f}_{0}(\bm{q}, \bm{\dot{q}}, t) + \bm{f}_{1}(\bm{q}, \bm{u}, t)$ &
    $\bm{0}_{n \times 1}$ & Kinematic differential equations \\
    $\bm{f}_{2}(\bm{q}, \bm{\dot{u}}, t) + \bm{f}_{3}(\bm{q}, \bm{\dot{q}},
    \bm{u}, \bm{r}, t)$ & $\bm{0}_{(o - m) \times 1}$ & Dynamic differential equations
  \end{tabular}
\end{table}

\begin{align}
\label{eq:quant_to_compute}
  \begin{array}{llcll}
    \tilde{M}_{qq}  &\triangleq \nabla_{\bm{\dot{q}}}\bm{f}_0 & \quad &
    \tilde{M}_{uqc} &\triangleq \nabla_{\bm{\dot{q}}}\bm{f}_a \\
    \tilde{M}_{uuc} &\triangleq \nabla_{\bm{\dot{u}}}\bm{f}_a & \quad &
    \tilde{M}_{uqd} &\triangleq \nabla_{\bm{\dot{q}}}\bm{f}_2 \\
    \tilde{M}_{uud} &\triangleq \nabla_{\bm{\dot{u}}}\bm{f}_2 & \quad &
    \tilde{A}_{qq}  &\triangleq -\nabla_{\bm{q}}(\bm{f}_0 + \bm{f}_1) \\
    \tilde{A}_{qu}  &\triangleq -\nabla_{\bm{u}}\bm{f}_1 & \quad &
    \tilde{A}_{uqc} &\triangleq - \nabla_{\bm{q}} \bm{f}_a \\
    \tilde{A}_{uuc} &\triangleq - \nabla_{\bm{u}} \bm{f}_a & \quad &
    \tilde{A}_{uqd} &\triangleq - \nabla_{\bm{q}} (\bm{f}_2 + \bm{f}_3) \\
    \tilde{A}_{uud} &\triangleq - \nabla_{\bm{u}} \bm{f}_3 & \quad &
    \tilde{B}_{u}   &\triangleq -\nabla_{\bm{r}}\bm{f}_{3}
  \end{array}
\end{align}

\begin{align}
  \label{eq:C_0}
  C_0 &\triangleq \left[ I_{n \times n} - P_{qd}(\nabla_{\bm{q}}
    \bm{f}_{c} P_{qd})^{-1} \nabla_{\bm{q}} \bm{f}_{c} \right] P_{qi}\\
  \label{eq:C_1}
  C_1 &\triangleq -P_{ud}(\nabla_{\bm{u}} \bm{f}_{v} P_{ud})^{-1}
  \nabla_{\bm{q}} \bm{f}_{v} \\
  \label{eq:C_2}
  C_2 &\triangleq \left[I - P_{ud} (\nabla_{\bm{u}}\bm{f}_{v} P_{ud})^{-1} \nabla_{\bm{u}}
    \bm{f}_{v} \right] P_{ui}\\
  \label{eq:Pq}
  \bm{q} &= P_{qi} \bm{q}_{i} + P_{qd} \bm{q}_{d} \\
  \label{eq:Pu}
  \bm{u} &= P_{ui} \bm{u}_{i} + P_{ud} \bm{u}_{d}
\end{align}

\section{Example}
\label{example}
We apply this linearization technique to the well studied rolling disc.  Figure
XX illustrates the choice of coordinates, speeds, and parameters.  The disc is
assumed to be knife edged of radius $r$, mass $m$, spin inertia $J$, and
transverse inertia $I$.  To illustrate the concept of dependent quantities, we
purposefully choose a nonminimal set of coordinates and speeds.  In this
particular example, this can be avoided, but in more complicated systems, it is
difficult (if not impossible) to describe the system with a minimal choice.  The
orientation of the disc fixed frame $C$ relative to the inertial frame $N$ is given by
a sequence of angular rotations (Euler ZXY) by angles $q_0$ (yaw), $q_1$
(lean), $q_2$ (spin).  The location
of the center of disc relative to the inertial origin is defined by coordinates
q3, q4, q5.  The angular velocity of the disc is defined by three speeds
\[u_0 \triangleq {^{N}}\bm{\omega}^{C} \cdot \bm{c}_{x}\]

@@

Equating the right hand sides of equations (\ref{eq:f0_plus_f1_expansion}),
(\ref{eq:acceleration_expansion}),
and (\ref{eq:f2_plus_f3_expansion}) to zero (as per Table
\ref{table:assumptions}), and introducing the following definitions
enables the unconstrained linear state space equations to be written as
\begin{align}
  \label{eq:state_space_unconstrained}
  \left[
    \begin{array}{cc}
      \tilde{M}_{qq} & \bm{0}_{n \times o} \\
      \tilde{M}_{uqc} & \tilde{M}_{uuc} \\
      \tilde{M}_{uqd} & \tilde{M}_{uud}
    \end{array}
    \right]
    \left[
      \begin{array}{c}
        \delta \bm{\dot{q}} \\
        \delta \bm{\dot{u}}
      \end{array}
    \right]
   &=
   \left[
     \begin{array}{cc}
       \tilde{A}_{qq} & \tilde{A}_{qu} \\
       \tilde{A}_{uqc} & \tilde{A}_{uuc} \\
       \tilde{A}_{uqd} & \tilde{A}_{uud}
     \end{array}
   \right]
    \left[
      \begin{array}{c}
        \delta \bm{q} \\
        \delta \bm{u}
      \end{array}
    \right]
    +
    \left[
      \begin{array}{c}
        \bm{0}_{(n + m) \times s} \\
        \tilde{B}_{u}
      \end{array}
    \right]
    \delta \bm{r}
\end{align}
Equation (\ref{eq:state_space_unconstrained}) has a state space of dimension $n
+ o$, yet only $p = n - l + o - m$ of these quantities are independent.
To address this issue, a particular set of independent
coordinates and speeds must be selected. To this end, consider the following
partitioning of the generalized coordinates and generalized speeds:
\begin{equation*}
  \tilde{\bm{q}} \triangleq \left[\begin{array}{cc}\bm{q}_{i} &
      \bm{q}_{d}\end{array}\right]^{T} =  P_{q}^{-1} \bm{q}
      \qquad\qquad
  \tilde{\bm{u}} \triangleq \left[\begin{array}{cc}\bm{u}_{i} &
      \bm{u}_{d}\end{array}\right]^{T} =  P_{u}^{-1} \bm{u}
\end{equation*}
where $P_q \in \mathbf{R}^{n \times n}$ and $P_u \in \mathbf{R}^{o \times o}$
are invertible permutation matrices which map an ordering which has the
independent quantities ($\bm{q}_{i}\in\mathbf{R}^{n-l},\,
\bm{u}_{i}\in\mathbf{R}^{o-m})$ first, followed by the dependent quantities
($\bm{q}_{d}\in\mathbf{R}^{l},\, \bm{u}_{d}\in\mathbf{R}^{m}$) to the
original ordering of the coordinates and speeds.  We use the notation $P_{qi}$
and $P_{qd}$ to denote the first $n-l$ and last $l$ columns of $P_q$,
respectively; similarly, $P_{ui}$ is the first $o-m$ columns of $P_{u}$ while
$P_{ud}$ is the last $m$ columns of $P_u$.

By assumption, equation
(\ref{eq:configuration_expansion}) is zero; making use of $P_q$, we have
\begin{align}
  \bm{0} &= \nabla_{\bm{q}}\bm{f}_{c} P_{q} \delta \bm{\tilde{q}} \notag \\
   &= \nabla_{\bm{q}}\bm{f}_{c} P_{qi} \delta \bm{q_i} +
  \nabla_{\bm{q}}\bm{f}_{c} P_{qd} \delta \bm{q_d}\notag\\
  \implies \delta \bm{q}_d &= -(\nabla_{\bm{q}}\bm{f}_{c} P_{qd})^{-1}
  (\nabla_{\bm{q}}\bm{f}_{c} P_{qi}) \delta \bm{q}_i \notag\\
  \implies \delta \bm{q} &= \left[ I_{n \times n} - P_{qd}(\nabla_{\bm{q}}
    \bm{f}_{c} P_{qd})^{-1} \nabla_{\bm{q}} \bm{f}_{c} \right] P_{qi} \delta
    \bm{q}_i
\end{align}
We define
which enables us to write
\begin{align}
  \label{eq:delta_q}
  \delta \bm{q} &= C_0 \delta \bm{q}_i
\end{align}

Applying the same approach to (\ref{eq:velocity_expansion}), we obtain
\begin{align}
  \bm{0} &= \nabla_{\bm{q}}\bm{f}_{v} \delta \bm{q} +
  \nabla_{\bm{u}}\bm{f}_{v} \delta \bm{u}\notag\\
  &= \nabla_{\bm{q}} \bm{f}_{v}
\delta \bm{q} + \nabla_{\bm{u}} \bm{f}_{v} P_{ui} \delta \bm{u}_i +
\nabla_{\bm{u}} \bm{f}_{v} P_{ud} \delta \bm{u}_d \notag\\
\implies \delta \bm{u}_d &= -\left(\nabla_{\bm{u}} \bm{f}_{v}
P_{ud}\right)^{-1}\left[\nabla_{\bm{q}}\bm{f}_{v} \delta\bm{q} +
  \nabla_{\bm{u}} \bm{f}_{v} P_{ui} \delta \bm{u}_i \right]\notag\\
  \implies \delta \bm{u} &= -P_{ud}(\nabla_{\bm{u}} \bm{f}_{v} P_{ud})^{-1}
  \nabla_{\bm{q}} \bm{f}_{v} \delta \bm{q}\notag\\
  &+ \left[I - P_{ud} (\nabla_{\bm{u}}\bm{f}_{v} P_{ud})^{-1} \nabla_{\bm{u}}
    \bm{f}_{v} \right] P_{ui} \delta \bm{u}_i
\end{align}
We define
which enables us to write
\begin{align}
  \label{eq:delta_u}
  \delta \bm{u} &= C_1 \delta \bm{q} + C_2 \delta \bm{u}_i\notag\\
  &= C_1 C_0 \delta \bm{q}_i + C_2 \delta \bm{u}_i
\end{align}

By making use of equations (\ref{eq:delta_q}) and (\ref{eq:delta_u}), we can
rewrite equation (\ref{eq:state_space_unconstrained}) as
%\begin{align}
%  \label{eq:state_space_constrained}
%  \left[
%    \begin{array}{cc}
%      \tilde{M}_{qq} & \bm{0}_{n \times o} \\
%      \tilde{M}_{uqc} & \tilde{M}_{uuc} \\
%      \tilde{M}_{uqd} & \tilde{M}_{uud}
%    \end{array}
%    \right]
%    \left[
%      \begin{array}{c}
%        \delta \bm{\dot{q}} \\
%        \delta \bm{\dot{u}}
%      \end{array}
%    \right]
%   &=
%   \left[
%     \begin{array}{cc}
%       (\tilde{A}_{qq} + \tilde{A}_{qu} C_1 ) C_0 & \tilde{A}_{qu} C_2 \\
%       (\tilde{A}_{uqc} + \tilde{A}_{uuc} C_1 ) C_0 & \tilde{A}_{uuc} C_2\\
%       (\tilde{A}_{uqd} + \tilde{A}_{uud} C_1 ) C_0 & \tilde{A}_{uud} C_2
%     \end{array}
%   \right]
%    \left[
%      \begin{array}{c}
%        \delta \bm{q}_i \\
%        \delta \bm{u}_i
%      \end{array}
%    \right]
%    +
%    \left[
%      \begin{array}{c}
%        \bm{0}_{(n+m) \times s} \\
%        \tilde{B}_{u}
%      \end{array}
%    \right]
%    \delta \bm{r}
%\end{align}
Equation \ref{eq:state_space_constrained} is a linear system of equations which
relates the time derivatives of all states (dependent and independent) to the
independent states and exogenous inputs; this is the precisely the form of
equation \ref{eq:structure}.
